Icicle: write once, run once

### 2016-05-30 2016fhpc/section/02-Source.tex
Pack down description of typing rules

Rules TcPrimApp and TcFunApp produce the type of the a primitive or function applied to its arguments.

% find the result type of primitive and function application.
% They first find the type of the primitive or function, and the types of the arguments, then use the application boxing/unboxing judgment $\TypecheckApp{\TauFun}{\ov{\TauMode}}{\TauMode}$ to find the return type and modality of the application.
% The set-comprehension syntax $\{e_i\}$ means that this rule allows any number of arguments.

Rule TcLet is a standard let rule, finding the type of the bound expression and inserting it into the context when typechecking the remaining expression.

Rule TcFold finds the type of a @fold@ expression.
The initial value of the fold must be pure.
The update has the current value of the fold inserted in its context as an @Element@, and the update must be an @Element@.
The final return type of the fold is an @Aggregate@.

Rule TcFilter requires its predicate to be a stream of booleans ($@Element@~\BB$).
The rest of the filter must be an @Aggregate@; returning an @Element@ stream with a different rate is not allowed, as explained in\sref{s:IcicleSource:bounded}.
Rule TcGroup is similar to filter, except returning a @Map@.

The rules PrimArith, PrimRel and PrimLookup give function types for each primitive.

The next judgment shows how function application boxes and unboxes modalities.
This judgment takes a function type and all its arguments, and returns the result type of the function.

Rule AppArgs applies when a function is applied to exactly the right arguments, and simply returns the result of the function.

Rule AppRebox is used when a pure function is applied to a non-pure value.
If the arguments and the result type are all pure, then the function can be applied to arguments of a particular modality, and the return type becomes that modality too.
Note that rule TcBox can also apply implicit boxing here; when only some of the arguments are non-pure, the pure ones can be boxed.
These rules are tricky to get right; an earlier version allowed applying a modal argument to a pure argument if the return type fitted that mode, which can lead to giving the results of folds as initial values for folds.

Rule CheckFun inserts the argument types into the context and returns a new context with the function's type, which is the type of the arguments to the result type.

Rule CheckQuery checks a query's types, and makes sure it returns an @Aggregate@.

Finally, rule CheckTop checks a whole top-level program.
It first creates a new context with the table columns bound to @Element@s, and typechecks each function.
Then, each query is checked, with the table columns and all functions available.
The final returned context is all query names and their types.

\subsection{Evaluation}

The evaluation rules for Icicle are given in figure~\ref{fig:source:eval}.

>>>>

Rules TcPrimApp and TcFunApp produce the type of the a primitive or function applied to its arguments. Rule TcLet is standard.

% find the result type of primitive and function application.
% They first find the type of the primitive or function, and the types of the arguments, then use the application boxing/unboxing judgment $\TypecheckApp{\TauFun}{\ov{\TauMode}}{\TauMode}$ to find the return type and modality of the application.
% The set-comprehension syntax $\{e_i\}$ means that this rule allows any number of arguments.
% Rule TcLet is a standard let rule, finding the type of the bound expression and inserting it into the context when typechecking the remaining expression.

In rule TcFold the initial value has value type $T$. The type of the current element from the stream is added to the context of $e_k$ as an @Element@, and the result of the overall fold is an @Aggregate@. Rules TcFilter and TcGroup are similar.

% The initial value of the fold must be pure.
% The update has the current value of the fold inserted in its context as an @Element@, and the update must be an @Element@.
% The final return type of the fold is an @Aggregate@.

% Rule TcFilter requires its predicate to be a stream of booleans ($@Element@~\BB$).
% The rest of the filter must be an @Aggregate@; returning an @Element@ stream with a different rate is not allowed, as explained in\sref{s:IcicleSource:bounded}.
% Rule TcGroup is similar to filter, except returning a @Map@.

The rules PrimArith, PrimRel and PrimLookup assign types to primitives. Rule AppArgs produces the type of a function or primitive applied to its arguments. Rule AppRebox is used when the arguments have modal type $m$ --- applying a pure function to arguments of mode $m$ produces a result of the same mode.

Rule CheckFun builds the type of a user defined function, returning it as an element of the output context. Rule CheckQuery is similar, noting that all queries return values of @Aggregate@ type. Finally, rule CheckTop checks a whole top-level program.

% The next judgment shows how function application boxes and unboxes modalities. This judgment takes a function type and all its arguments, and returns the result type of the function.

% Rule AppArgs applies when a function is applied to exactly the right arguments, and simply returns the result of the function.

% Rule AppRebox is used when a pure function is applied to a non-pure value.
% If the arguments and the result type are all pure, then the function can be applied to arguments of a particular modality, and the return type becomes that modality too.
% Note that rule TcBox can also apply implicit boxing here; when only some of the arguments are non-pure, the pure ones can be boxed.
% These rules are tricky to get right; an earlier version allowed applying a modal argument to a pure argument if the return type fitted that mode, which can lead to giving the results of folds as initial values for folds.

% inserts the argument types into the context and returns a new context with the function's type, which is the type of the arguments to the result type.

% Rule CheckQuery checks a query's types, and makes sure it returns an @Aggregate@.

% It first creates a new context with the table columns bound to @Element@s, and typechecks each function.
% Then, each query is checked, with the table columns and all functions available.
% The final returned context is all query names and their types.

% ---------------------------------------------------------
\subsection{Evaluation}
The evaluation rules for Icicle are given in figure~\ref{fig:source:eval}.
